---
title: 'Two-Sample Inference Procedures for Population Central Values'
author: "Solutions"
output: word_document
fontsize: 12pt
---

```{r include=FALSE, cache=FALSE}
# Don't modify this chunk of code, it is just installing and loading the DS705data package
if (!require(DS705data)){
  if (!require(devtools)){
    install.packages('devtools',repos="http://cran.rstudio.com")
  }
  library(devtools)
  install_github('DataScienceUWL/DS705data')
}
require(DS705data)
```
   
## Exercise 1  

An experiment was conducted to evaluate the effectiveness of a drug treatment for tapeworm in the stomachs of sheep. A random sample of 24 worm-infected lambs of approximately the same age was randomly divided into two groups. Twelve of the lambs were injected with the drug and the remaining twelve were left untreated. After a 6-month period the lambs were slaughtered and the worm counts recorded.

### Part 1a

Load the dataset WormSheep from the DS705data package. Note that the package is loaded above at line 18.  You just need the data() command.  To see all the data sets in the package, type data(package='DS705data').

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
data(WormSheep)
```

---

### Part 1b	

Create boxplots, histograms, and normal probability plots for each group (treated vs. untreated).  Be sure that each plot is labeled appropriately.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1b -|-|-|-|-|-|-|-|-|-|-|-

```{r, fig.width=3, fig.height=3}
boxplot(worms~treated,data=WormSheep)
wormsT <- WormSheep$worms[WormSheep$treated=="treated"]
wormsUT <- WormSheep$worms[WormSheep$treated=="untreated"]
hist(wormsT,main="Treated Sheep",xlab="Worms")
hist(wormsUT,main="Treated Sheep",xlab="Worms")
qqnorm(wormsT)
qqline(wormsT)
qqnorm(wormsUT)
qqline(wormsUT)
```

---

### Part 1c  

Do the boxplots show any outliers? 

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1c -|-|-|-|-|-|-|-|-|-|-|-

There are no outliers indicated on the boxplot.

---

### Part 1d  

Describe the shapes of the histograms for the sample data for each sample.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1d -|-|-|-|-|-|-|-|-|-|-|-

They are roughly mound-shaped and the worm counts for the untreated sheep may be a little skewed to the right.  However, with such small samples we don't really have enough data to establish a strong pattern.

---

### Part 1d  

Do either of the normal probability plots show evidence of non-normality?  Explain.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1d -|-|-|-|-|-|-|-|-|-|-|-

Neither of the normal probability plots show evidence of non-normality because the dots do not deviate much from the diagonal line in each plot.

---

### Part 1f 

Conduct an appropriate test to determine if the worm counts in each population can be considered as normally distributed.  Provide the p-value and the conclusion of the test.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1f -|-|-|-|-|-|-|-|-|-|-|-

```{r}
shapiro.test(wormsT)
shapiro.test(wormsUT)
pT <- shapiro.test(wormsT)$p.value
pUT <- shapiro.test(wormsUT)$p.value
```

For untreated sheep the Shapiro-Wilk test of the worm count data gives a P-value of `r round(pT,3)`.  Since $P$ is large, do not reject the null hypothesis that the worm counts are normally distributed.  For treated sheep, the P-value is `r round(pUT,3)` and the conclusion is the same as for the untreated sheep.


### Part 1g	
    
Conduct the test of your choice to determine if the population mean worm count for all sheep treated with the drug is less than the mean worm count for the population of untreated sheep.  Let $\alpha = 0.05$.

#### Step 1

State the null and alternative hypotheses for the test.  If you use symbols, please define them.
 
#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1g.step1 -|-|-|-|-|-|-|-|-|-|-|-

If using symbols, define them as follows:

$\mu_{\mbox{treated}}$ = population mean worm count for treated sheep.

$\mu_{\mbox{untreated}}$ = population mean worm count for untreated sheep.

You can typeset hypotheses like this (using LaTeX):
$$H_0: \mu_\mbox{treated} = \mu_\mbox{untreated}$$
$$H_a: \mu_\mbox{treated} < \mu_\mbox{untreated}$$


#### Step 2

Use R to generate the output for the test you selected.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1g.step2 -|-|-|-|-|-|-|-|-|-|-|-

```{r}
test <- t.test(worms~treated,data=WormSheep,alternative="less")
test
```

#### Step 3

State both a statistical conclusion at $\alpha=0.05$ and interpret it in the context of the problem.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 1g.step3 -|-|-|-|-|-|-|-|-|-|-|-

Reject $H_0$ at $\alpha = 0.05$. There is sufficient evidence to suggest that the population mean worm counts are different for the treated and untreated sheep ($P =$ `r round(test$p.value,3)`).

---

### Part 1h  

Compute a 95% confidence interval for the difference in population means.  Write an interpretation in the context of the problem for this interval.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1h -|-|-|-|-|-|-|-|-|-|-|-

```{r}
ci <- t.test(worms~treated,data=WormSheep,alternative="two.sided")
ci
```

We are 95% confident that the population mean worm count for treated sheep is `r abs(round(ci$conf.int[2],2))` to `r abs(round(ci$conf.int[1],2))` less than the population mean worm count for untreated sheep.

---

### Part 1j  

Did you use the Welch t-test or the Wilcoxon rank sum test?  Justify your choice, including some discussion of how well the conditions for the hypothesis test and confidence interval procedures were met.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1j -|-|-|-|-|-|-|-|-|-|-|-

The histograms are roughly mound shaped and the Shapiro-Wilk test indicates both samples could be from normally distributed populations.  Taken together, these reasons suggest the Welch t-procedures are most appropriate.  

---

## Problem 2

Consider two independent, random samples of delay times (in hour) for two airlines over the past 10 years. The task here is to compare the population central values by conducting a hypothesis test and constructing a 95% confidence interval for the difference in average delay time between the airlines for the population of all delays in the past 10 years.

### Part 2a  

Load the dataset delays from the DS705data package.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
data(delays)
```

---

### Part 2b

Begin by exploring the sample means, medians, and standard deviations for each airline. That is, find the sample means, medians, and standard deviations. 

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
print("Means")
with(delays,tapply(delay,airline,mean))
print("Medians")
with(delays,tapply(delay,airline,median))
print("Standard deviations")
with(delays,tapply(delay,airline,sd))
```

---

### Part 2b

Produce boxplots, histograms, and normal probability plots for the delay times of both airlines.  

Discuss the aspects of the plots that are relevant to the construction of the best type of test and confidence interval to compare the population central values.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
boxplot(delay~airline,data=delays,xlab="Delay Time",horizontal=T)
attach(delays)
hist(delay[airline=="Skybird"])
hist(delay[airline=="Omega"])
qqnorm(delay[airline=="Skybird"])
qqline(delay[airline=="Skybird"])
qqnorm(delay[airline=="Omega"])
qqline(delay[airline=="Omega"])
detach(delays)
```


Both distributions are very skewed to the right with numerous outliers on the upper end, but generally of the same shape.  The Wilcoxon Rank Sum test with the Hodges-Lehmann interval (which is the kind produced by wilcox.test) is most appropriate.

---

### Part 2c

Conduct hypothesis tests for the normality of the delay times for both Skybird and Omega using a 5% level of significance in each case.

Use R to compute the test statistic and P-value.  

State the hypotheses and conclusion of the test. Include the P-value in your conclusion.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2c -|-|-|-|-|-|-|-|-|-|-|-

```{r}
sky <- shapiro.test(delay[airline=="Skybird"])
omega <- shapiro.test(delay[airline=="Omega"])
```

In each case,

$H_0$: The sample comes from a normally distributed population.

$H_a$: The sample does not come from a normally distributed population.

For Skybird, reject $H_0$ at $\alpha=0.05$.  The population of delay times for all Skybird delays is not normally distributed ($P =$ `r round(sky$p.value,4)`).

For Omega, reject $H_0$ at $\alpha=0.05$.  The population of delay times for all Omega delays is not normally distributed ($P =$ `r round(omega$p.value,4)`).
 
---

### Part 2d	

Perform the most appropriate test to compare population central values for all Skybird and Omega delay times for the past 10 years.  Use a 10% level of significance.

#### Step 1

State the null and alternative hypotheses for the test (in words).
 
#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2d.step1 -|-|-|-|-|-|-|-|-|-|-|-

$H_0$: The distributions of delay times over the past 10 years for Skybird and Omega are identical.

$H_a$: The distributions of delay times over the past 10 years for Skybird and Omega are shifted from each other.


#### Step 2

Use R to generate the output for the test you selected.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2d.step2 -|-|-|-|-|-|-|-|-|-|-|-

```{r}
wtest <- wilcox.test(delay~airline,data=delays,conf.int=T)
wtest
```

#### Step 3

State both a statistical conclusion at $\alpha=0.05$ and interpret it in the context of the problem.

#### -|-|-|-|-|-|-|-|-|-|-|- Answer 2d.step3 -|-|-|-|-|-|-|-|-|-|-|-

Do not reject $H_0$ at $\alpha = 0.05$. There is insufficient evidence to conclude that the distributions of delay times over the past 10 years for Skybird and Omega are shifted from each other ($P =$ `r round(wtest$p.value,3)`).

---

### Part 2e

Construct the most appropriate 95% confidence interval to compare population central values.  Write a sentence to interpret the interval in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2e -|-|-|-|-|-|-|-|-|-|-|-

```{r}
wtest <- wilcox.test(delay~airline,data=delays,conf.int=T)
wtest$conf.int
```

With 95% confidence, the difference in population median delay times for Omega and Skybird airlines, with the order of subtraction being Diff = Omega - Skybird, is between `r round(wtest$conf.int[1],2)` and `r round(wtest$conf.int[2],2)`.

---

### Part 2f

Did you use the Welch t-test and interval or the Wilcoxon rank sum test with the corresponding interval?  Justify your choice, including some discussion of how well the conditions for the hypothesis test and confidence interval procedures were met.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2f -|-|-|-|-|-|-|-|-|-|-|-

The samples were said to be independent and randomly selected in the problem statement, so those data conditions are met.  Since both distributions were quite skewed with many outliers, but were roughly the same shape, the Wilcoxon rank sum test with the Hodges-Lehman confidence interval for the difference in medians were used.

---

### Part 2g

What type of hypothesis testing error could have been made here?  Describe it in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2g -|-|-|-|-|-|-|-|-|-|-|-

Since the null hypothesis was not rejected, a Type II error might have occurred, meaning that the populations of flight delays for Skybird and Omega really are shifted from each other, but the sample was such that the test didn't reveal it.

---

### Part 2h

Construct a 95% bootstrap confidence interval to compare population medians for the delay times over the past 10 years for the two airlines. Use the Bca method and a bootstrap sample size of 2000. Compare it to the interval produced by the wilcox.test function.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2h -|-|-|-|-|-|-|-|-|-|-|-

```{r}
bootPctDiff <- function(d,i,p=.5){
  pct <- tapply(d[i,1],d[,2],quantile,prob=p)
  pct[1]-pct[2]
  }

boot.object <- boot(delays, bootPctDiff, R = 2000,p=.5,
                    strata=delays$airline)
bootci <- boot.ci(boot.object,type='bca')$bca[4:5]
bootci
```

The 95% bootstrap confidence interval from `r round(bootci[1],2)` to `r round(bootci[2],2)` and is similar to the Hodges-Lehmann interval produced by wilcox.test.  In each case the interval covers 0 and so does not provide enough evidence to claim the population medians differ.



