
---
title: "DS 705 Final Exam"
author: "Matt Allen"
output:
  word_document: default
  pdf_document: default
fontsize: 12pt
---

Unlike the Homework RMD files, this one doesn't contain all of the questions.  Rather we want you to read the questions on D2L and use this RMD file to record your R code that you used to answer each question.  This file must knit correctly.  After you submit the D2L quiz, then upload this RMD file and the knitted version of this file to the Final Exam dropbox on D2L.

## Questions 1-16

You do not need to submit any work with these questions.  Just answer them in D2L.

## Problem 1 - Questions 17-21

You're going to analyze the data in BirdPecks.rda and answer the questions in D2l.  Put all of your R in the chunk below:

```{r}
load(file = "BirdPecks.rda")
boxplot(BirdPecks$pecks~BirdPecks$group)
hist(BirdPecks$pecks[which(BirdPecks$group==1)], main="Pecks Group 1", xlab = "Pecks")
hist(BirdPecks$pecks[which(BirdPecks$group==2)], main="Pecks Group 2", xlab = "Pecks") 
#test for normality
shapiro.test(BirdPecks$pecks[which(BirdPecks$group==1)])
shapiro.test(BirdPecks$pecks[which(BirdPecks$group==2)])
#Wilcoxon rank sum test
wtest <- wilcox.test(pecks~group,data=BirdPecks,conf.int=T,conf.level = 0.90)
wtest
```

---

## Problem 2 - Questions 22-26

Analyze StrengthSpeed.rda and put your R below:

```{r}
load(file = "StrengthSpeed.rda")
StrengthSpeed$diff <- StrengthSpeed$after-StrengthSpeed$before
#create plots
boxplot(StrengthSpeed$diff)
hist(StrengthSpeed$diff)
{
  qqnorm(StrengthSpeed$diff)
  qqline(StrengthSpeed$diff)
}
shapiro.test(StrengthSpeed$diff)
#perform paired t-test
test <-t.test(StrengthSpeed$diff,mu=0,alternative="less",conf.level=0.95)
test
```

---

## Problem 3 - Questions 27-36

Analyze GroupHLT scores and put your R here:

```{r}
hlt.after.win.1 <- c(79, 76, 74, 70, 81, 85, 73, 78, 69, 72, 83, 89, 72, 79, 75)
group1 <- rep(1, length(hlt.after.win.1))
hlt.after.loss.2 <- c(78, 96, 85, 91, 77,103, 72, 93, 98, 86, 70, 110, 70, 91, 99)
group2 <- rep(2, length(hlt.after.loss.2))

hostility.scores <- data.frame(group=c(group1,group2), hlt.score=c(hlt.after.win.1,hlt.after.loss.2))

boxplot(hlt.score~group,data=hostility.scores)
hist(hlt.after.win.1,main="Hostility After Win",xlab="HLT Score")
hist(hlt.after.loss.2,main="Hostility After Loss",xlab="HLT Score")
{
  qqnorm(hlt.after.win.1)
  qqline(hlt.after.win.1)
}
{
  qqnorm(hlt.after.loss.2)
  qqline(hlt.after.loss.2)
}
#do shapiro-wilk test
shapiro.test(hlt.after.win.1)$p.value
shapiro.test(hlt.after.loss.2)$p.value

#perform test
test <- t.test(hlt.score~group,data=hostility.scores,alternative="less")
test

#confidence interval after loss
t.test(hlt.after.loss.2, conf.level = 0.95)

#bootstrap confidence interval after loss
library(boot)
samplemean <- function(x, d) {
  return(mean(x[d]))
}
boot.object <- boot(hlt.after.loss.2, samplemean, R = 5000)
boot.ci(boot.object,conf=.95,type='bca')$bca[4:5]
```

---

## Problem 4 - Questions 37-41

Analyze the data in treadware.rda and put your R here:

```{r}
#load(file = "treadware.rda") #this command is failing with error:Error in readChar(con, 5L, useBytes = TRUE) : cannot open the connection
#use ds705data instead
require(DS705data)
data("treadwear")
aggregate(wear~brand,data=treadwear, FUN= "mean" )
aggregate(wear~brand,data=treadwear, FUN= "sd" )
boxplot(wear~brand,data=treadwear)

shapiro.test(treadwear[treadwear$brand == 'A',]$wear)$p.value
shapiro.test(treadwear[treadwear$brand == 'B',]$wear)$p.value
shapiro.test(treadwear[treadwear$brand == 'C',]$wear)$p.value
shapiro.test(treadwear[treadwear$brand == 'D',]$wear)$p.value

#multiple comparisons procedure
#Games-Howell
onewayComp(wear~brand,data=treadwear,var.equal=FALSE,
           adjust='one.step')$pair[[3]]<.05
```

---

## Problem 5 - Questions 42-59

Analyze the data in diamond.rda and include your R here:

```{r}
load(file = "diamond.rda")
plot(diamond$price~diamond$carat)
#linear model
linear.model <- lm(price~carat,data=diamond)
summary(linear.model)
#pearson correlation coefficient
out <- cor.test(diamond$carat,diamond$price)
out
rPearson <- out$estimate
rPearson
ci <- out$conf.int
ci

#confidence interval linear model coefficients
confint(linear.model)

#normality of residuals
resids <- linear.model$resid
hist(resids)
{
qqnorm(resids)
qqline(resids)
}
shapiro.test(resids)

price.fit <- linear.model$fitted.values
plot(price.fit,resids)

#test for equal variance
require(lmtest)
out <- bptest(linear.model)
p <- out$p.value
p
#coefficient of determination
summary(linear.model)
```

---

## Problem 6 - Questions 60- 66

Analyze the data in Shells.rda and include your R here:

```{r}
load(file = "shells.rda")

fit1 <- lm(Y~.,data=shells)
fit.null <- lm(Y~1,data=shells)

step(fit.null,scope=list(lower=fit.null,upper=fit1),direction="both")
#check for multicollinearity
#install.packages("car")
library(car)
model.a <- lm(formula = Y ~ X2 + X4 + X1 + X6 + X7, data = shells)
vif(model.a)
summary(model.a)
#model b
model.b <- lm(formula = Y ~ X1+X2+X2**2+X4+X6, data = shells)
model.b
extractAIC(model.b)
summary(model.b)
```

---

## Problem 7 - Questions 67-70

Analyze the "Primary News Source for Americans" data described in the problem statement.  Put your R below:

```{r}
observed <- c(38,20,15,42)
proportions <- c(.45,.18,.16,.21)
chisq.test(x=observed,p=proportions)
```

---

## Problem 8 - Questions 71-75

Analyze the data in cheese.rda and put your R below:

```{r}
load(file = "cheese.rda")
summary(cheese)
boxplot(acetic~taste, data=cheese)
#fit logistic model
cheese.out <- glm(taste~acetic + person,data=cheese,family="binomial")
summary(cheese.out)
#odds.ratio acetic
exp(2.787)

newdata <- data.frame(person="Child", acetic=6)
predict(cheese.out, newdata, type="response")

#conf int
out <- predict(cheese.out, newdata,se.fit=TRUE)
C = .95  # define the level of confidence
crit = qnorm(1-(1-C)/2)  # get the appropriate critical value
lower = exp(out$fit-crit*out$se.fit)/(1+exp(out$fit-crit*out$se.fit))
upper = exp(out$fit+crit*out$se.fit)/(1+exp(out$fit+crit*out$se.fit))
c(lower,upper)
```

---

## Problem 9 - Questions 76-90

Analyze the data in careerbarrier.rda and put your R below:

```{r}
load(file = "careerbarrier.rda")

require(psych)
mat <- cor(careerbarrier[,1:15])
cortest.bartlett(mat,n=76)

#KMO MSA
KMO(mat)

#drop money
mat <- cor(careerbarrier[,2:15])
cortest.bartlett(mat,n=76)

#KMO MSA
KMO(mat)


#scree plot
output <- princomp(careerbarrier[,2:15], cor=TRUE)
{
plot(output,type="lines") # scree plot 
abline(h=1,lty=2)
}
output$sdev^2  # This prints the eigenvalues so you can compare to 1

#Use a principal components extraction with the varimax rotation to extract 5 factors
fan <- principal(careerbarrier[,2:15],nfactors=5,rotate="varimax")
print(fan,cut=.5,sort=TRUE)

```

---

## Problem 10 - Questions 91-98

Analyze the data on seat postion vs. nausea described in the problem.  Put your R below:

```{r}
nausea <-c(98,110,161)
no.nausea<-c(264,321,280)

seatposition <- matrix(c(nausea,no.nausea),ncol=3,byrow=TRUE)
colnames(seatposition) <- c("Front","Middle","Rear")
rownames(seatposition) <- c("Nausea","No Nausea")
seatposition <- as.table(seatposition)
seatposition

addmargins(seatposition)

result <- chisq.test(seatposition)
result

nausea.bus.frontvsrear=c(98,161)
Total=c(362,441)
out <- prop.test(nausea.bus.frontvsrear,Total,correct=FALSE, conf.level = 0.9) # continuity correction is optional
out

#odds ratio
(161/280)/(98/264)

```

---

## Question 99

Make sure both this RMD and the resulting knitted Word document are uploaded to the Dropbox "Final Exam R Code."
